From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-30 WOLFHPC15/article/mscac.tex
overview helene 29_07
introduction helene 29_07
overview fix helene 30_07

As illustrated in Figure~\ref{fig:mscac}, the compiler, called MSCAC for \emph{Multi-Stencil Component Assembly Compiler}, is composed of three different phases. The first phase of the compiler is to \emph{parse} the MSL description. Second, the compiler analyzes and transforms the computation part of the description, in other words $\Gamma$, to an intermediate representation. This phase of compilation, called MSC, is responsible for the automatic parallelization of the overall numerical simulation. Finally, the parallel intermediate representation can be dumped to an actual implementation of the parallel structure of the simulation.

@@

%In this paper are presented the language MSL, the MSC compilation phase, and the dump compilation phase. 
Parallelization techniques handled by MSC are \emph{data parallelism}, where data are splitted among processors or cores while applying a single program everywhere, and \emph{task parallelism} where the program is transformed into a dependency graph of tasks. From a dependency graph different strategies can be used to schedule tasks, in other words to control their execution in a valid order. The MSC compiler presented in this paper handles the transformation of the dependency graph to a static schedule of the tasks.

The final phase of compilation, the dump, uses the static schedule of the tasks, identified from $\Gamma$, and additionnal information on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{D}_m$ and $\mathcal{D}_c$, from the \emph{parser}, to produce a back-end solution which represents the parallel pattern or skeleton of the overall simulation. We could have propose a dump to OpenMP~\cite{660313}, however this paper presents a first contribution to dump a DSL to a component-based runtime. Component models, where a component represents a clear and independent functionnality of the application, have first been proposed in Software engineering and have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. Those properties, on the other hand, are recurrent stated problems of parallel applications.

The back-end of the compiler is to build a parallel assembly of components, in other words appropriate connections between functionnalities of the numerical simulation. This assembly is illustrated in Figure~\ref{fig:assembly} which is explained in details in Section~\ref{sect:component}. As illustrated in the figure, each component is linked to one or more tuple of the multi-stencil program definition: \emph{DDS} component represents $\mathcal{M}$ and $\mathcal{D}_m$; \emph{Data} component represents $\Delta$ and $\mathcal{D}_c$; \emph{Time} component represents $T$, and \emph{Computations} component (which is generated by MSC) represents $\Gamma$.

The final back-end component assembly is a ready-to-fill parallel structure where the user has to write computation codes described in the MSL description. To write those computations, interfaces of the choosen implementation have to be used. For example, the back-end evaluated in this paper uses the distributed data structure proposed in SkelGIS library~\cite{CPE:CPE3494}, however a work in progress replaces this DDS component by another one using Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} data structure.

@@

   \node[provide,below=0.15 of DAu2] (Tp) {};
   \node[component,below=0.8cm of Tp] (T) {$Time$};
   \node[use,right=1cm of T] (Tu) {};

   \node[provide,right=0.15 of Tu] (Cp) {};
   \node[component,right=2cm of Cp] (C) {$Computations$};
   \node[use,above=0.8cm of C] (Cu) {};
   \node[right=0.2cm of Cu] (star) {$*$};

   \node[provide,right=0.2 of DAu1] (Datap1) {};
   \node[provide,above=0.15 of Cu] (Datap2) {};
   \node[component,above=0.8cm of Datap2] (Data) {$Data$};
   \node[use,above=0.8cm of Data] (Datau) {};

   \node[provide,right=0.15 of Du1] (DDSp1) {};
   \node[provide,above=0.15 of Datau] (DDSp2) {};
   \node[component,above=0.8cm of DDSp2] (DDS) {$DDS$};

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
    (DDSp1) edge node {} (DDS)
    (DA) edge node {} (DAu1)
           edge node {} (DAu2)
    (DAp) edge node {} (DA)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\caption{Shape of the generated component assembly.}
\label{fig:assembly}
\end{center}
\end{figure}

>>>>

As illustrated in Figure~\ref{fig:mscac}, the compiler, called MSCAC for \emph{Multi-Stencil Component Assembly Compiler}, is composed of three different phases. The first phase of the compiler is to \emph{parse} the MSL description. Second, the compiler analyzes and transforms the computation part of the description $\Gamma$, to an intermediate representation. This phase of compilation, called MSC, is responsible for the automatic parallelization of the overall numerical simulation. Finally, the parallel intermediate representation is dumped to an actual implementation of the parallel pattern of the simulation.

@@

%In this paper are presented the language MSL, the MSC compilation phase, and the dump compilation phase. 
Parallelization techniques handled by MSC are \emph{data parallelism}, where data are splitted among processors or cores while applying a single program everywhere, and \emph{task parallelism} where the program is transformed into a dependency graph of tasks. From a dependency graph different strategies can be used to schedule tasks, in other words to control their execution in a valid order. The MSC compiler presented in this paper handles the transformation of the dependency graph to a static scheduling.

The final phase of compilation, the dump, uses the static schedule of the tasks, created from $\Gamma$, and additionnal information on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{D}_m$ and $\mathcal{D}_c$, given by the \emph{parser}, to produce a back-end solution which represents the parallel pattern or skeleton of the overall simulation. We could have propose a dump to OpenMP~\cite{660313}, however this paper presents a first contribution to dump a DSL to a component-based runtime. Component models, where a component represents a clear and independent functionnality of the application, have first been proposed in Software engineering and have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. Those properties, on the other hand, are recurrent stated problems of parallel applications.

The back-end of the compiler is to build a parallel assembly of components, in other words the appropriate connections between functionnalities of the numerical simulation. This assembly is illustrated in Figure~\ref{fig:assembly} which is explained in details in Section~\ref{sect:component}. As illustrated in the figure, each component is linked to one or more tuple of the multi-stencil program definition: \emph{DDS} component represents $\mathcal{M}$ and $\mathcal{D}_m$; \emph{Data} component represents $\Delta$ and $\mathcal{D}_c$; \emph{Time} component represents $T$, and \emph{Computations} component (which is generated by MSC) represents $\Gamma$.

The final back-end component assembly is a ready-to-fill parallel structure where the user has to write computation codes described in the MSL description. To write those computations, interfaces of the choosen implementation have to be used. For example, the back-end evaluated in this paper uses the distributed data structure proposed in SkelGIS library~\cite{CPE:CPE3494}. A work in progress replaces this DDS component by another one using Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} data structure.

@@

   \node[provide,below=0.15 of DAu2] (Tp) {};
   \node[component,below=0.8cm of Tp] (T) {$Time$};
   \node[use,right=1cm of T] (Tu) {};
   \node[left=0.8cm of T] (tt) {$T$};

   \node[provide,right=0.15 of Tu] (Cp) {};
   \node[component,right=2cm of Cp] (C) {$Computations$};
   \node[use,above=0.8cm of C] (Cu) {};
   \node[right=0.2cm of Cu] (star) {$*$};
   \node[right=1.5cm of C] (gamma) {$\Gamma$};

   \node[provide,right=0.2 of DAu1] (Datap1) {};
   \node[provide,above=0.15 of Cu] (Datap2) {};
   \node[component,above=0.8cm of Datap2] (Data) {$Data$};
   \node[use,above=0.8cm of Data] (Datau) {};
   \node[right=1cm of Data] (delta) {$\Delta,\mathcal{D}_c$};

   \node[provide,right=0.15 of Du1] (DDSp1) {};
   \node[provide,above=0.15 of Datau] (DDSp2) {};
   \node[component,above=0.8cm of DDSp2] (DDS) {$DDS$};
   \node[right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{D}_m$};

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
    (DDSp1) edge node {} (DDS)
    (DA) edge node {} (DAu1)
           edge node {} (DAu2)
    (DAp) edge node {} (DA)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\caption{Overall back-end component assembly}
\label{fig:assembly}
\end{center}
\end{figure}
